\documentclass[a4paper]{article}

\input{prelude.tex}

\title{Quantum Computing Meets Type Theory:\\
 \Large A Proposal for a Framework for Certified Quantum Information Processing
% \\ \LARGE (Justification of Resources)
}
\author{(Justification of Resources) --- Thorsten Altenkirch and Alexander S. Green}
\date{}

\begin{document}

\twocolumn[\maketitle]{

\section*{Personnel}
\label{sec:personnel}

The research suggested in the proposal is based on the PhD of
Alexander S. Green supervised by Thorsten Altenkirch. 
Altenkirch as the lead investigator will spend 15\% of his time on the
project, on which Green would be a full time RA. A PhD student working on
one particular work package will support the project, and this will
also contribute to the education of young scientists i the area of
quantum computing and type theory. 

Since the project involves both theoretical work and the
implementation of a software prototype, a running time of 4 years is
justified. Moreover, in our experience, PhD projects in an innovative
and theoretical area usually take at least four years. Hence by
applying for a four year project it is possible for the student to
finish his PhD during the running time of the project,

\subsection*{Conference attendance}
\label{sec:travel}

We plan to attend conferences and workshops relevant to our research,
with the aim to learn about progress in the area and to dissiminate
our own results. Topical workshops are often most productive, but it
is often difficult to predict exactly when and how often they will
take place. We would attend meetings on quantum computing, and in
particular meetings on the interaction of quantum information
processing and logic (e.g. QPL (Quantum Physics and Logic), QUOXIC
(Quantum meetings of Imperial College and Oxford); the
Reversible Computation workshop, the workshops on Categories, Logic
and Foundations of Physics), meetings related to dependently typed
programming (e.g. TYPES workshops, DTP (Workshop on Dependently Typed
Programming; AIM (the Agda Implementers Meeting); the Coq workshop);
meetings on functional programming (e.g. TFP (Trends in Functional
Programming) ; ICFP (the International Conference on Functional
Programming)) and general Computer Science conferences, especially if
they include a session on quantum computing (e.g. ( CiE (Computability
in Europe), POPL (Principles of Programming Languages); LICS (Logic in
Computer Science)).

We estimate that the PI and the RA will attend 1 international and 1-2
UK/EU workshops or conferences each year, and that the PhD student
will attend 2 summer schools, 2 national/EU and 1 international
workshop or conference. We estimate the costs for attending an
international conference as \pounds 1500, and national/EU conferences
and summer school as \pounds 1000. Thus, over the running time of
the project we would have 9 international and 14 national/EU
conference visits and 2 summer school attendances.

\subsection*{Collaboration}
\label{sec:collaboration}

We plan to collaborate closely with Lucas Dixon in Edinburgh and Ross
Duncan, and with Samson Abramsky and Bob Coecke in Oxford. This
collaboration is important to obtain early feedback on the usefulness
of our design (WP1) and specifically on using high level structures
(WP4), i.e. Coecke's and Abdamsky's categorical models and their
implementation in Dixon's and Duncan's \emph{Quantomatics}.  We envisage
5 visits lasting approximately one week, and budget \pounds 800 each
for travel (appr. \pounds 150) , accomodation (appr. \pounds 400) and subsistence
(appr. \pounds 250), i.e. \pounds 4000 over the lifetime of the project.

% We are also planning to collaboarate with collegues and such as Coecke
% and Abramsky at Oxford, Selinger at Dalhousie, Panagedan at McGill,
% \dots. Hence we include visits by the PI and the RA lasting 1-2 weeks
% each, we estimate that there will be 1-2 visits among all the
% participants a year at a cost of \pounds 1200. Over the running time
% of the project there would be 6 visits.

% \subsection*{Workshop and Spring School}
% \label{sec:worksh-summ-scho}

% We will organize a workshop on \textit{Formal Methods in Quantum Computing} 
% in Nottingham during the middle of the project to discuss
% our work with other researchers but also to obtain valuable input on
% our work. We would like to fund travel
% and accommodation (est. $\pounds 1000$ each) for two invited speakers.

% We also plan to present a course at the Midland Graduate School and
% include \pounds 800 to cover accomodation and travel for this.

\section*{Equipment and consumables}
\label{sec:equipment}

The research proposed here  necessitates the development of experimental
prototype software, seldom optimized at first, and requiring
frequent time-consuming recompilation. Correspondingly, we
request funding for high spec desktop PCs and laptops
to support these investigations effectively.

We will equip each of the participants with a laptop computer
(est. cost $\pounds 1500$) for presentations and work while not at the
office.  Laptops are essential for joint work at the miniworkshops and
during visits and are not supplied by the university.
The researcher and the doctoral student will 
require a high spec desktop computer (at $\pounds 1000$ each).

\end{document}

